---
title: "Welcome to the jungle"
description: | 
  I have reluctantly decided to create a new blog. Some thoughts on 
  what I hope to achieve, having tried my hand at blogging so very many times 
  before
params:
  slug: welcome
author:
  - first_name: "Danielle"
    last_name: "Navarro"
    url: https://djnavarro.net
    affiliation: UNSW Sydney
    affiliation_url: https://unsw.edu.au
    orcid_id: 0000-0001-7648-6578
date: 2021-04-05
creative_commons: CC BY
output:
  distill::distill_article:
    self_contained: false
---


<!--------------- my typical setup ----------------->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r box, echo=FALSE}
# import utility functions as a module
box::use(../../source/utils[
  set_meta, 
  set_redirect, 
  set_source, 
  set_session,
  set_lockfile,
  set_timestamp
])
```

```{r meta, echo=FALSE}
set_meta(
  title =  "Welcome to the jungle", 
  description = "Some thoughts on starting over, when I have tried my hand at blogging many times before"
)
```


<!--------------- post begins here ----------------->

I've decided the time has come to restart my blog. I've tried blogging many times before with mixed success, and this time around I'd like to avoid the mistakes of the past. I've set up this blog with a few principles in mind:

- **Simplicity**. One mistake I've often made is to create blogs using the fanciest tools I could find. For example, I've previously used Hugo based packages like blogdown and hugodown, and much as I love those tools (and use them on other sites) I want this blog to be as low-maintenance as possible. To that end I'm using distill, and I'm keeping the default settings in most respects.

- **Encapsulation**. There was a time when I really liked the idea of having my blog integrated nicely with my homepage ([djnavarro.net](https://djnavarro.net)). I've become less keen on this because the aesthetic and technical demands of a blog aren't always aligned with the needs of my homepage. This time I've set it up so that the  [blog.djnavarro.net](https://blog.djnavarro.net) subdomain corresponds to a different repository from my homepage. I'm hoping this will make blogging simpler from a technical standpoint.

- **Focus**. Another mistake I have made in the past is letting blogs "sprawl", mixing personal essays with technical posts. My intention with this blog is to write technical posts only, mostly on R and data science. I've moved my personal writing to [essays.djnavarro.net](https://essays.djnavarro.net) and my artwork to [art.djnavarro.net](https://art.djnavarro.net). My hope is that this will make blogging easier from an emotional standpoint. 

- **Reproducibility**. A frustration I've had with my previous blogs is that my posts were not particularly reproducible. Source code was often missing, information about the R session was nor provided, and so on. This time, I've set up the blog so that there is a "details" section at the bottom of each post containing links to the source code, the R session information, and a lockfile generated using `renv::snapshot()`.

Not sure how this will go, but fingers crossed! `r emo::ji("crossed_fingers")`



<!--------------- appendices go here ----------------->

## Last updated {.appendix}

`r set_timestamp("Australia/Sydney")`

## Details {.appendix}

[source code](`r set_source(params$slug)`), [session info](`r set_session(params$slug)`), [lockfile](`r set_lockfile(params$slug)`)


<!--------------- miscellanea ----------------->

```{r redirect, echo=FALSE}
set_redirect(params$slug)
```


