---
title: "Bootstrap cards in distill"
description: |
  How to enable bootstrap 4 on a distill website, even though you probably 
  don't need to. I like it though because I get to add pretty bootstrap cards
params:
  slug: bs4cards-in-distill
author:
  - first_name: "Danielle"
    last_name: "Navarro"
    url: https://djnavarro.net
    affiliation: UNSW Sydney
    affiliation_url: https://unsw.edu.au
    orcid_id: 0000-0001-7648-6578
date: 04-19-2021
bibliography: bs4cards-in-distill.bib
creative_commons: CC BY
citation_url: https://blog.djnavarro.net/bs4cards-in-distill
repository_url: https://github.com/djnavarro/distill-blog
output:
  distill::distill_article:
    self_contained: false
    toc: true
    css: "strapless.css"
---


<!--------------- my typical setup ----------------->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r box, echo=FALSE}
# ensure box knows where this post is located
box::set_script_path(
  path = here::here("_posts", params$slug, paste0(params$slug, ".Rmd"))
)

# import utility functions as a module
box::use(../../source/utils[
  set_meta, 
  set_redirect, 
  set_source, 
  set_session,
  set_lockfile,
  set_timestamp
])
```

```{r meta, echo=FALSE}
set_meta(
  title =  "Bootstrap cards in distill", 
  description = "How to enable bootstrap 4 cards on a distill website"
)
```


<!--------------- post begins here ----------------->


When creating R markdown websites, I often find myself wanting to organise content into a nice-looking grid of links. For example, in a [recent project](https://slides.djnavarro.net/) I wanted to be able to create something like this:

```{r, echo=FALSE}
library(bs4cards)
slides <- list(
  
  card(
    title = "Starting R markdown",
    image = "https://slides.djnavarro.net/index_img/starting-rmarkdown.png",
    link = "https://slides.djnavarro.net/starting-rmarkdown",
    text = "An introduction to R markdown. The target audience is a novice R user with no previous experience with markdown."
  ),
  
  card(
    title = "Starting ggplot2",
    image = "https://slides.djnavarro.net/index_img/starting-ggplot2.png",
    link = "https://slides.djnavarro.net/starting-ggplot2",
    text = "An introduction to ggplot2. The target audience is a novice user with no previous experience with R or ggplot2."
  ),
  
  card(
    title = "Starting programming",
    image = "https://slides.djnavarro.net/index_img/starting-programming.png",
    link = "https://slides.djnavarro.net/starting-programming",
    text = "This is primarily a tutorial on making generative art in R, but in doing so introduces core programming constructs and data structures."
  )
)

card_grid(slides)
```

It bothered me that this wasn't as straightforward as I was expecting, so for one of my side projects I've been putting together a small package called [bs4cards](https://bs4cards.djnavarro.net) to make this a little easier inside an R markdown document or website. There's an [introductory vignette](https://bs4cards.djnavarro.net/articles/introduction.html) posted on the bs4cards package website showing how the package works, and there's no need to duplicate that content here. However, because this website uses the [distill](https://rstudio.github.io/distill/) package [@distill] and the package website is built using [pkgdown](https://pkgdown.r-lib.org/) [@pkgdown], it seems like a good idea to have at least one post on both sites that uses bs4cards. 


## Enabling bootstrap 4

The reason for doing this is that the first step in using the package is to make sure that your R markdown document uses version 4 of bootstrap: the `bs4cards` package takes its name from the [cards](https://getbootstrap.com/docs/4.5/components/card/) system introduced in bootstrap version 4, and will not work properly if used in R markdown documents that rely on bootstrap version 3, or don't use bootstrap at all. To ensure that you are using bootstrap 4, you need to edit the YAML header for your document to specify which version of bootstrap you want to use. The instructions are slightly different depending on what kind of document you're creating:

### Vanilla R markdown

For a plain R markdown document or website (i.e., one where the output format is `html_document`) here is the relevant section of YAML you might use:

```
output:
  html_document:
    theme:
      version: 4
```

This overrides the R markdown defaults [@rmarkdown] to ensure that the output is built using [bootstrap 4.5](https://getbootstrap.com/docs/4.5/getting-started/introduction/). 

### Pkgdown

To enable bootstrap 4 in a pkgdown site, the process is similar but not identical. Edit the `_pkgdown.yml` file to include the following

```
template:
  bootstrap: 4
```

Note that this relies on a currently-in-development feature, so you may need to update to the development version of pkgdown to make this work.


### Distill

Distill R markdown does not use bootstrap, which is a little inconvenient if you want to use bs4cards with distill. With a little effort it is possible to enable the entire bootstrap library in a distill site, but this can lead to undesirable side-effects because bootstrap has a lot of styling that doesn't look visually appealing when mixed with the distill styling.^[The original version of this post used that solution, with a promissory note to do a better job later.] To fix this, bs4cards includes a stripped down version of bootstrap called "strapless", that only uses the subset of bootstrap css relevant to cards. The convenience function `use_strapless()` will copy the strapless css file to the target location. To copy it to the same location as this post:

```{r, echo=TRUE}
bs4cards::use_strapless(here::here("_posts", "bs4cards-in-distill"))
```

Next, to enable the library within the post, add the following lines to the YAML header for the post:

```
output:
  distill::distill_article:
    css: "strapless.css"
```

Alternatively, if you want to enable strapless for the entire site, this markup goes in the `_site.yml` file and the css file should go in the home folder for the project.

Once that's done you should be ready to go! However, please be careful: this is an experimental feature and has not yet been properly tested.

## Testing with kittens

Just to make certain, let's check that it does what we want by generating a random card using kitten images and lorem ipsum placeholder text:

```{r, echo=TRUE}
bs4cards::sample_card()
```

<br>

Looks about right to me?



<!--------------- appendices go here ----------------->

## Last updated {.appendix}

`r set_timestamp("Australia/Sydney")`

## Details {.appendix}

[source code](`r set_source(params$slug)`), [session info](`r set_session(params$slug)`), [lockfile](`r set_lockfile(params$slug)`)


<!--------------- miscellanea ----------------->

```{r redirect, echo=FALSE}
set_redirect(params$slug)
```


