---
title: "Title in sentence case" # <---- UPDATE ME
description:
  Neque porro quisquam est qui dolorem ipsum quia dolor sit amet, 
  consectetur, adipisci velit # <---- UPDATE ME
author:
  - first_name: "Danielle"
    last_name: "Navarro"
    url: https://djnavarro.net
    affiliation: UNSW Sydney
    affiliation_url: https://unsw.edu.au
    orcid_id: 0000-0001-7648-6578
preview: preview-image.jpg  # <---- UPDATE ME 
creative_commons: CC BY
output:
  distill::distill_article:
    self_contained: false
params:
  slug: tidy-tuesday-billboard
  date: 2021-09-14
  repo: djnavarro/distill-blog
  site: https://blog.djnavarro.net/
  data: https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-14/billboard.csv
---

<!----

checklist:
  - check the "update me" messages in YAML above
  - initialise the _renv folder with refinery::renv_new(long_slug)
  - populate the lockfile with refinery::renv_snapshot(long_slug)
  - update the _renv folder from snapshot with refinery::refresh(long_slug)

---->


<!--------------- setup post ----------------->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
refinery::renv_load(paste(params$date, params$slug, sep = "_"))
```

<!--------------- post ----------------->

I've never participated in [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday) before, but since I've joined a slack that does, it is high time I did soemthing about that poor track record. 

The data are located [here](`r params$data`). I've stored the link in the YAML header for this post so I can read the data with the following code, and set `cache=TRUE` in the chunk options to avoid downloading the data each time I knit the post again:

```{r import-data, cache=TRUE}
billboard <- readr::read_csv(params$data)
billboard
```

At vero eos et accusamus et iusto odio dignissimos ducimus qui blanditiis praesentium voluptatum deleniti atque corrupti quos dolores et quas molestias excepturi sint occaecati cupiditate non provident, similique sunt in culpa qui officia deserunt mollitia animi, id est laborum et dolorum fuga. Et harum quidem rerum facilis est et expedita distinctio. Nam libero tempore, cum soluta nobis est eligendi optio cumque nihil impedit quo minus id quod maxime placeat facere possimus, omnis voluptas assumenda est, omnis dolor repellendus. Temporibus autem quibusdam et aut officiis debitis aut rerum necessitatibus saepe eveniet ut et voluptates repudiandae sint et molestiae non recusandae. Itaque earum rerum hic tenetur a sapiente delectus, ut aut reiciendis voluptatibus maiores alias consequatur aut perferendis doloribus asperiores repellat.


<!--------------- appendices ----------------->

```{r, echo=FALSE}
refinery::insert_appendix(
  repo_spec = params$repo, 
  dir = paste(params$date, params$slug, sep = "_")
)
```


<!--------------- miscellanea ----------------->

```{r redirect, echo=FALSE}
refinery::insert_netlify_redirect(
  slug = params$slug, 
  date = params$date
)
```


<!--------------- derived YAML ----------------->
---
date: `r params$date`
citation_url: `r paste0(params$site, params$slug)`
repository_url: `r paste0("https://github.com/", params$repo)` 
---



