---
title: "Generative art, jasmines, and the water colours series"
description: | 
  Neque porro quisquam est qui dolorem ipsum quia dolor sit amet, 
  consectetur, adipisci velit
params:
  slug: generating-the-water-colours
  date: 2021-09-07
author:
  - first_name: "Danielle"
    last_name: "Navarro"
    url: https://djnavarro.net
    affiliation: UNSW Sydney
    affiliation_url: https://unsw.edu.au
    orcid_id: 0000-0001-7648-6578
date: 2021-09-07
preview: jasmine-recollected.png
creative_commons: CC BY
citation_url: https://blog.djnavarro.net/generating-the-water-colours
repository_url: https://github.com/djnavarro/distill-blog
output:
  distill::distill_article:
    toc: true
    css: "strapless.css"
    self_contained: false
---

<!----

yaml checklist:
- set the preview image
- write title and description

renv checklist:
- initialise the _renv folder with refinery::renv_new(long_slug)
- populate the lockfile with refinery::renv_snapshot(long_slug)
- update the _renv folder from snapshot with refinery::refresh(long_slug)

setup, appendices, and redirects should "just work" (ish)

---->

<!--------------- setup ----------------->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width="100%")
long_slug <- paste(params$date, params$slug, sep = "_")
# refinery::renv_load(long_slug) # <-- disable for now
```

```{r environment, include=FALSE, message=FALSE}
library(magrittr)
library(readr)
library(tidyr)
library(tibble)
library(stringr)
library(ggplot2)
library(purrr)
library(dplyr)

conflicted::conflict_prefer("extract", "tidyr")
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
```


<!--------------- post ----------------->

> I find it so amazing when people tell me that electronic music has no soul. You can't blame the computer. If there's no soul in the music, it's because nobody put it there ([Björk](https://www.azquotes.com/quote/564301), via [Tim de Sousa](https://twitter.com/TimdeSousa/status/1432898952907554822))


## Prelude

In recent weeks I've been posting generative art from the [Water Colours](https://art.djnavarro.net/gallery/water-colours/) series on twitter. The series has been popular, prompting requests that I sell prints, mint NFTs, or write a tutorial showing how they are made. For personal reasons I didn't want to commercialise this series. Instead, I chose to make the pieces freely available under a CC0 public domain licence and asked people to donate to a [gofundme](https://au.gofundme.com/f/data-science-art-to-support-a-womens-refuge) I set up for a charitable organisation I care about (the [Lou's Place](https://www.lousplace.com.au/) women's refuge here in Sydney). I'm not going to discuss the personal story behind this series, but it does matter. As I've mentioned [previously](/posts/2021-07-08_generative-art-in-r/), the art I make is inherently tied to moods and emotions. In hindsight it is easy enough to describe how the system is implemented but this perspective is misleading. Although a clean and unemotional description of the code is useful for explanatory purposes, the actual *process* of creating the system is deeply tied to my life, my history, and my subjective experience. Those details are inextricably bound to the system. A friend described it better than I ever could:

> The computer doesn't make this art any more than a camera makes a photograph; art is always intimate ([Amy Patterson](https://twitter.com/justanotheramy/status/1432916731794579460))

In this post I'll describe the mechanistic processes involved in creating these pieces, but this is woefully inadequate as a description of the artistic process as a whole. The optical mechanics of a camera do not circumscribe the work of a skilled photographer. So it goes with generative art. The code describes the mechanics; it does not describe the art. There is a deeply personal story underneath these pieces, and I would no more mint an NFT from that story than I would sell a piece of my soul to a collector.

## The water colours repository

When I started making generative art I didn't think much about archiving my art or keeping it organised. I liked making pretty things, and that was as far as my thought process went. I didn't place the code under version control, and I stored everything in my Dropbox folder. There's nothing wrong with that: some things don't belong on GitHub. During the development phase of any art project that's still what I do, and I'm perfectly happy with it.

Things become a little trickier when you want to share the art. My [art website](https://art.djnavarro.net) is hosted on GitHub pages, and so my initial approach was to keep the art in the website repository. *Huuuuge mistake.* Sometimes the image files can be quite large and sometimes a series contains a large number of images. By the time I'd reached 40+ series, Hugo took a very long time to build the site (several minutes), and GitHub took even longer to deploy it (over half an hour).

Eventually I decided it made more sense to have one repository per series. Each one uses the "series-" prefix to remind me it's an art repo. I don't use these repositories during development: they exist solely to snapshot the release. For example, the [series-water-colours](https://github.com/djnavarro/series-water-colours) repository isn't going to be updated regularly, it's really just an archive combined with a "docs" folder that is used to host a minimal GitHub Pages site that makes the images public. It's convenient for my purposes because my art website doesn't have to host any of the images: all it does is hotlink to the images that are exposed via the series repo.

It may seem surprising that I've used GitHub for this. Image files aren't exactly well suited to version control, but it's not like they're going to be updated. Plus, there are a lot of advantages. I can explicitly include licencing information in the repository, I can release source code (when I want to), and I can include a readme file for anyone who wants to use it. 

One nice feature of doing things this way is that it has encouraged me to include a [manifest file]("https://djnavarro.net/series-water-colours/manifest.csv"), a plain csv file with one row per image, and one column for each piece of metadata I want to retain about the images. Again, it seems like overkill, but I'm starting to realise that if I don't have a system I'll forget things like "what the piece is called" or "when I made it". 

```{r water-colour-manifest, echo=FALSE, message=FALSE, cache=TRUE}
manifest <- read_csv("https://djnavarro.net/series-water-colours/manifest.csv")

manifest <- manifest %>% 
  filter(
    resolution %in% c(500, 2000), 
    format == "jpg"
  ) %>% 
  mutate(
    path = paste0("https://djnavarro.net/series-water-colours/", path)
  ) %>% 
  pivot_wider(
    names_from = resolution, 
    names_prefix = "path_", 
    values_from = path
  )
```

I can use `readr::read_csv()` to download the manifest and do a little data wrangling to organise it into a format that is handy to me right now:

```{r show-manifest}
manifest
```

The data wrangling I used to get it into this format isn't shown in the post, but you can always look at the [source code](https://github.com/djnavarro/distill-blog/tree/master/_posts/2021-09-07_generating-the-water-colours) for this post if you're curious. More to the point, the `manifest` data frame is nicely suited for use with the [bs4cards](https://bs4cards.djnavarro.net/) package, so I can display some of the pieces in a neat and tidy thumbnail grid:

```{r water-colour-cards, message=FALSE, cache=TRUE}
manifest[1:8, ] %>% 
  bs4cards::cards(
    image = path_500,
    link = path_2000,
    title = long_name,
    spacing = 3,
    width = 2
  )  
```

Each thumbnail image links to a medium resolution (2000 x 2000 pixels) jpg version of the corresponding piece, if you'd like to see the images in a little more detail. 

## Dependencies 

A subset of tidyverse packages are loaded 

```{r}
library(magrittr)
library(readr)
library(tidyr)
library(tibble)
library(stringr)
library(ggplot2)
library(purrr)
library(dplyr)
```

Besides tidyverse and base R functions, I'll use a few other packages as well: in addition to the bs4cards package that made a brief appearance earlier, the [magick](https://docs.ropensci.org/magick/), [raster](https://github.com/rspatial/raster/), [rprojroot](https://rprojroot.r-lib.org/), [fs](https://fs.r-lib.org/), and [ambient](https://ambient.data-imaginist.com/) packages are all used in making the art. Because functions from those packages may not be as familiar to you, I'll namespace the calls to them in the same way I did with `bs4cards::cards()` previously. Hopefully that will make it easier to see which functions belong to one of those packages


## Art by image processing

As in life, the place to start is knowing where you are. 

```{r define-paths}
name <- paste(params$date, params$slug, sep = "_")
blog <- rprojroot::find_root("_site.yml")
post <- fs::path(blog, "_posts", name)
file <- fs::path(post, "jasmine.jpg")
```

Here it is:

```{r jasmine-photo}
knitr::include_graphics(file)
```

The photo has an emotional resonance to me: it dates back to 2011 and appeared on the cover of [Learning Statistics with R](https://learningstatisticswithr.com). In fact, although 10 years separate the *Water Colours* series from the text book and the jasmines photo, the two are linked by a shared connection to events from a decade ago. Nor is it coincidental that "Jasmine" is my middle name. 


The next step is importing the image, scaling it down to a manageable resolution, and then converting it to matrix. The magick and raster packages are doing the work here:

```{r import-image}
import_image <- function(path, width, height) {
  geometry <- paste0(width, "x", height)
  path %>% 
    magick::image_read() %>% 
    magick::image_scale(geometry)
}
```

```{r construct-matrix}
construct_matrix <- function(image) {
  
  # read matrix
  mat <- image %>% 
    as.raster() %>%
    as.matrix()
  
  # use the row and column names to represent co-ordinates
  rownames(mat) <- paste0("y", nrow(mat):1) # <- flip y
  colnames(mat) <- paste0("x", 1:ncol(mat))
  
  return(mat)
}
```

```{r construct-tibble}
construct_tibble <- function(mat) {
  
  # convert to tibble
  tbl <- mat %>%
    as.data.frame() %>%
    rownames_to_column("y") %>%
    as_tibble() 
  
  # reshape
  tbl <- tbl %>%
    pivot_longer(
      cols = starts_with("x"),
      names_to = "x",
      values_to = "shade"
    ) 
  
  # tidy
  tbl <- tbl %>%
    arrange(x, y) %>% 
    mutate(
      x = x %>% str_remove_all("x") %>% as.numeric(),
      y = y %>% str_remove_all("y") %>% as.numeric(),
      id = row_number()
    )
  
  return(tbl)
}
```

```{r jasmines-data}
jas <- file %>% 
  import_image(width = 100, height = 60) %>% 
  construct_matrix() %>% 
  construct_tibble()

jas
```

Every plot is going to use this, so:

```{r ggplot-themed}
ggplot_themed <- function(data) {
  data %>% 
    ggplot(aes(x, y)) +
    coord_equal() + 
    scale_size_identity() + 
    scale_colour_identity() + 
    scale_fill_identity() + 
    theme_void() 
}
```

We can use `geom_tile()` to draw the original image:

```{r jasmine-raster-image}
jas %>% 
  ggplot_themed() + 
  geom_tile(aes(fill = shade)) 
```

As an aside, if you map the borders of each cell to the shade you get something neat:

```{r jasmine-raster-variant-image}
jas %>% 
  ggplot_themed() + 
  geom_tile(aes(colour = shade)) 
```

I didn't go down that path originally, but it's worth noting that there might be something worth playing with. Generative art, much like any other kind of art, is as much about exploration and discovery as it is about technical prowess. 

Corresponding scatter plot:

```{r jasmine-scatter-image}
jas %>% 
  ggplot_themed() + 
  geom_point(aes(colour = shade)) 
```

Next step, extract the channels

```{r extract-channels}
extract_channels <- function(tbl) {
  rgb <- with(tbl, col2rgb(shade))
  hsv <- rgb2hsv(rgb)
  tbl <- tbl %>% 
    mutate(
      red = rgb[1, ],
      grn = rgb[2, ],
      blu = rgb[3, ],
      hue = hsv[1, ],
      sat = hsv[2, ],
      val = hsv[3, ]
    )
  return(tbl)
}
```

```{r jasmine-channels-data}
jas <- extract_channels(jas)
jas
```  

One way to use this representation is in halftone images. If you have a printer that contains only black ink, you can approximate shades of grey by using the size of each dot to represent how dark that pixel should be:

```{r jasmine-halftone-image}
map_size <- function(x) {
  ambient::normalise(1-x, to = c(0, 2))
}

jas %>% 
  ggplot_themed() +  
  geom_point(
    mapping = aes(size = map_size(val)),
    colour = "black", 
    show.legend = FALSE
  )
```

For real world printers, this approach is very convenient because it allows us to construct any shade we like using only a few different colours of ink. In the halftone world shades of grey are merely blacks of different size, pinks are merely sizes of red (sort of), and so on. But we're not using real printers. There's nothing stopping us retaining the original hue and saturation, while using dot size to represent intensity. That allows us to produce "halftonesque" images:


```{r jasmine-halftonesque-image}
jas %>% 
  ggplot_themed() +  
  geom_point(
    mapping = aes(
      colour = hsv(hue, sat, .5), 
      size = map_size(val)
    ), 
    show.legend = FALSE
  )
```


## Intermission

```{r water-colour-cards-2, message=FALSE, echo=FALSE, cache=TRUE}
manifest[9:12, ] %>% 
  bs4cards::cards(
    image = path_500,
    link = path_2000,
    title = long_name,
    spacing = 3,
    width = 2
  )  
```


## Art from noise generators


Generating vector fields using the ambient package: 

```{r field}
field <- function(points, frequency = .1, octaves = 1) {
  ambient::curl_noise(
    generator = ambient::fracture,
    fractal = ambient::billow,
    noise = ambient::gen_simplex,
    x = points$x,
    y = points$y,
    frequency = frequency,
    octaves = octaves,
    seed = 1
  )
}
```

```{r shift}
shift <- function(points, amount, ...) {
  vectors <- field(points, ...)
  points <- points %>%
    mutate(
      x = x + vectors$x * amount,
      y = y + vectors$y * amount,
      time = time + 1,
      id = id
    )
  return(points)
}
```

```{r noise-stepped-data}
points_time0 <- expand_grid(x = 1:50, y = 1:30) %>% 
  mutate(time = 0, id = row_number())

points_time1 <- shift(points_time0, amount = 1)
points_time2 <- shift(points_time1, amount = 1)
points_time3 <- shift(points_time2, amount = 1)

pts <- bind_rows(
  points_time0, 
  points_time1, 
  points_time2,
  points_time3
)
```

```{r noise-stepped-image}
map_size <- function(x) {
  ambient::normalise(x, to = c(0, 2))
}

map_alpha <- function(x) {
  ambient::normalise(-x, to = c(0, .5))
}

pts %>% 
  ggplot_themed() +  
  geom_point(
    mapping = aes(
      size = map_size(time), 
      alpha = map_alpha(time)
    ),
    show.legend = FALSE
  )
```

The purrr package provides a useful tool for iterating:

```{r iterate}
iterate <- function(pts, time, step, ...) {
  bind_rows(accumulate(
    .x = rep(step, time), 
    .f = shift, 
    .init = pts,
    ...
  ))
}
```

```{r extract-points}
extract_points <- function(data) {
  data %>% 
    select(x, y, id) %>% 
    mutate(time = 0)
}
```

```{r accumulated-noise-data}
pts <- jas %>% 
  extract_points() %>% 
  iterate(time = 20, step = .1)
```

```{r accumulated-noise-image, cache=TRUE}
map_size <- function(x) {
  ambient::normalise(x^2, to = c(0, 5))
}

pts %>% 
  ggplot_themed() +  
  geom_point(
    mapping = aes(size = map_size(time)),
    alpha = .01,
    show.legend = FALSE
  ) 
```


## Assembling the parts

```{r join-parts}
join_parts <- function(jas, pts) {
  jas %>% 
    select(-x, -y) %>% 
    full_join(pts, by = "id") %>% 
    arrange(time, id) 
}

```

```{r noisy-jasmine-data, cache=TRUE}
jas <- join_parts(jas, pts)
jas
```

```{r noisy-jasmine-image, cache=TRUE}
map_size <- function(x, y) {
  ambient::normalise((1 - x) * y^2, to = c(0, 5))
}

jas %>% 
  ggplot_themed() +  
  geom_point(
    mapping = aes(
      colour = hsv(hue, sat, .5), 
      size = map_size(val, time)
    ), 
    alpha = .03,
    show.legend = FALSE
  )
```


Repeat at higher resolution:

```{r jasmine-detailed-data, cache=TRUE}
jas <- file %>% 
  import_image(width = 200, height = 120) %>% 
  construct_matrix() %>% 
  construct_tibble() %>% 
  extract_channels()

pts <- jas %>% 
  extract_points() %>% 
  iterate(
    time = 40, 
    step = .2, 
    octaves = 10, 
    frequency = .05
  )

jas <- join_parts(jas, pts)
```


```{r jasmine-detailed-image, cache=TRUE}
map_size <- function(x, y) {
  12 * (1 - x) * (max(y)^2 - y^2) / y^2
}

pic <- jas %>% 
  ggplot_themed() +  
  geom_point(
    mapping = aes(
      colour = shade, 
      size = map_size(val, time)
    ), 
    alpha = 1,
    stroke = 0,
    show.legend = FALSE
  ) 

pic
```


```{r jasmine-cropped-image, cache=TRUE}
pic +
  scale_x_continuous(limits = c(11, 190), expand = c(0, 0)) +
  scale_y_continuous(limits = c(7, 114), expand = c(0, 0))
```


## Epilogue

```{r water-colour-cards-3, message=FALSE, echo=FALSE, cache=TRUE}
manifest[13:20, ] %>% 
  bs4cards::cards(
    image = path_500,
    link = path_2000,
    title = long_name,
    spacing = 3,
    width = 2
  )  
```


<!--------------- appendices ----------------->

```{r, echo=FALSE}
refinery::insert_appendix("djnavarro/distill-blog", long_slug)
```


<!--------------- miscellanea ----------------->

```{r redirect, echo=FALSE}
refinery::insert_netlify_redirect(params$slug, params$date)
```


