---
title: "Pretty little CLIs"
description: |
  How to make a gorgeous command line interface in R using the cli package.
params:
  slug: pretty-little-clis
author:
  - first_name: "Danielle"
    last_name: "Navarro"
    url: https://djnavarro.net
    affiliation: UNSW Sydney
    affiliation_url: https://unsw.edu.au
    orcid_id: 0000-0001-7648-6578
date: 04-18-2021
bibliography: pretty-little-clis.bib
creative_commons: CC BY
citation_url: https://blog.djnavarro.net/pretty-little-clis
repository_url: https://github.com/djnavarro/distill-blog
output:
  distill::distill_article:
    toc: true
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)

# uncomment this line to generate the asciicast svgs
# source(here::here("_posts", "pretty-little-clis", "source", "create-asciicasts.R"))

# convenience function to insert an asciicast svg
insert_asciicast <- function(name) {
  dir <- here::here("_posts", "pretty-little-clis", "output")
  svg <- paste0(name, ".svg")
  knitr::include_graphics(file.path(dir, svg))
}
```


```{r display-secrets, echo=FALSE}
insert_asciicast("secrets")
```

<aside>
Lyrics to the title theme of the US TV show, [*Pretty Little Liars*](https://prettylittleliars.fandom.com/). The song is called *Secrets*, taken from the fabulous [Thirteen Tales of Love and Revenge](https://en.wikipedia.org/wiki/Thirteen_Tales_of_Love_and_Revenge) album by *The Pierces*
</aside>

Anytime you write R code whose output needs to be understood by a human being, it is an act of kindness to spend a little time making sure that the output shown to the human being properly communicates with that human. As a consequence of this, you often find yourself needing to write information to the R console, just to cater to those precious human sensibilities. Perhaps the simplest way to do this is to use the `cat()` function. It's a simple tool and it gets the job done in most cases. If the mysterious "A" from *Pretty Little Liars* had used R to craft her threatening text messages, she could have written code like this:

```{r cat-example, eval=FALSE}
wait <- function(seconds = 2) {
  Sys.sleep(seconds)
}

send_cat_threat <- function() {
  cat("Dead girls walking.\n"); wait()
  cat("--A.\n")
}
```

Equipped with a function that specifies her threat, complete with a dramatic pause for effect, she's ready to go. When her unwitting victim does something to trigger the `send_cat_threat()` function, a two part message is displayed on the console. The first part shows up immediately

```{r cat-example-partial, echo=FALSE}
cat("Dead girls walking.")
```

and after a two second delay, her call sign is revealed

```{r, ref.label="cat-example", echo=FALSE, eval=TRUE}
```

It's not too difficult to imagine what As message might look like at the R console, but where's the fun in that? Thanks to the [asciicast](https://github.com/r-lib/asciicast) package [@asciiicast], there's no need to leave anything to the imagination, and we can see the malevolent message in screencast form:

```{r display-cat-example, echo=FALSE}
insert_asciicast("cat-example")
```

<aside>
The ominous text messages used in this post are taken from *Pretty Little Liars*. 
</aside>

Using `cat()` to craft messages works perfectly well for simple text communication, but sometimes you want something that looks a little fancier. After all, if the big picture plan here is to impersonate a dead teenager and terrorise her friends - and for some reason you've chosen R to do so - you might as well put a little effort into the details, right?

## Meet the cli package

One thing I love about the R community is that if you search long enough you'll find that someone else has already written a wonderful package that solves the problem you're facing. If your problem is "how to craft nicely formatted messages" then you'll be delighted to learn that many wonderful things become possible if you have the [cli](https://cli.r-lib.org) package [@cli] as your talented assistant. To craft a beautiful command line interface (CLI) of our very own, the first thing we'll need to do is load the package:

```{r package-load}
library(cli)
```

Once this is done, it is a very trivial task for A to write the same threatening text message using `cli_text()`...

```{r cli-example-1, eval=FALSE}
send_cli_threat <- function() {
  cli_text("Dead girls walking."); wait()
  cli_text("--A.")
}
send_cli_threat()
```
```{r display-cli-example-1, echo=FALSE}
insert_asciicast("cli-example-1")
```

...which is nice and all, but it doesn't make much of a case for using cli. Stalking and threatening is busy work, and I'd imagine that A would want a more compelling justification before deciding to switch her evil workflow. However, much liky A herself, the R console has many dark secrets, and fancier tricks than this are possible once you know how to expose them using cli.

## Using the status bar

One piece of magic that I have wondered about for a long time is how fancy progress bars work: often when you're doing something that takes a long time, you'll see a progress bar rendered in text on the screen, which suddenly vanishes once the process is complete. While teaching myself cli, I found the answer. The most recent line of text generated at the terminal is called the *status bar*: the state of the status bar can be manipulated, and the cli package provides a neat toolkit for doing so. So let's say I were trying to convince A to switch to the cli tools. Right now, she's writing a function that will send a four-part message, using `cli_text()` because I've at least convinced her to *try* the new tools:

```{r status-bar-1, eval=FALSE}
message_scroll <- function() {
  cli_text("You found my bracelet."); wait()
  cli_text("Now come find me."); wait()
  cli_text("Good luck bitches."); wait()
  cli_text("-A"); wait()
}
message_scroll()
```

When her victim triggers this message the lines will appear on screen, one after the other with an appropriate dramatic pause separating them. The victim might see something that looks like this:

```{r display-status-bar-1, echo=FALSE}
insert_asciicast("status-bar-1")
```

The problem -- when viewed from an evil point of view -- is that this message stays on screen after delivery.^[Yes, it does disappear in this screencast, but that's just the screencast. If it were the R console it would stay on screen the whole time.] The victim has time to think about it, take a screenshot to show her friends, that kind of thing. Wouldn't the gaslighting be so much more effective if she were to send the message piece by piece, each part disappearing as the next one appears, only to have the whole thing vanish without a trace and leaving the victim wondering if she imagined the whole thing? This is where the status bar comes in handy. Here's how it would work:


```{r status-bar-2-function, eval=FALSE}
message_inline <- function() {
  id <- cli_status("")
  cli_status_update(id, "You found my bracelet."); wait()
  cli_status_update(id, "Now come find me."); wait()
  cli_status_update(id, "Good luck bitches."); wait()
  cli_status_update(id, "-A"); wait()
  cli_status_clear(id)
}
```

The first line in this function uses `cli_status()` to create a blank message on the status bar, and returns an identifier that refers to the status bar. The next four lines all use `cli_status_update()` to overwrite the current state of the status bar, and then pause dramatically for two seconds. In a final act of malice, the last line in the function clears the status bar using `cli_status_clear()`, leaving nothing except a blank space behind. So what the victim sees is something more like this:

```{r status-bar-2, eval=FALSE}
message_inline()
```
```{r display-status-bar-2, echo=FALSE}
insert_asciicast("status-bar-2")
```


## Creating spinners

The ability to control the status bar opens up a world of new possibilities. Progress bars are one such possibility, but the [progress](https://github.com/r-lib/progress) package [@progress] already does this nicely, and in any case I suspect that A might be more intrigued by the possibility of spinners, since they just spin and spin and give the victim no clue about when the process is going to end. Much more appealing when the developer doesn't know (or doesn't want to reveal) when the wait will end. The cli package has a nice `makes_spinner` function that serves this purpose. Here's an example:

```{r spinner-0, eval=FALSE} 
spinny <- make_spinner(
  which = "dots2",
  template = "{spin} It's not over until I say it is."
)
```

The `which` argument is used to choose how the spinner would look, and the `template` argument is used to define how the "spinny bit" is placed relative to the rest of the text. The spinny object includes functions to update the state of the spinner (in this case `spinny$spin()` would be that function), and a function to clear the spinner from the status bar. So here's how A might define a function that uses a spinner to keep the victim in suspense...


```{r spinner-1, eval=FALSE}
theatrics <- function(which) {
  
  # define the spinner
  spinny <- make_spinner(
    which = which,
    template = "{spin} It's not over until I say it is."
  )
  
  # update the spinner 100 times
  for(i in 1:100) {
    spinny$spin()
    wait(.05)
  }
  
  # clear the spinner from the status bar
  spinny$finish()
  
  # send the final part of the message
  cli_alert_success("Sleep tight while you still can, bitches. -A")
}
```

Here's what happens:

```{r spinner-1-call, eval=FALSE}
theatrics("dots2")
```
```{r display-spinner1, echo=FALSE}
insert_asciicast("spinner-1")
```

Setting `which = "dots2"` is only one possibility. There are quite a lot of different spinner types that come bundled with the cli package, and I'd imagine A would want to look around to see which one suits her needs. Personally, I'm a fan of hearts:

```{r spinner-2, eval=FALSE}
theatrics("hearts")
```
```{r display-spinner2, echo=FALSE}
insert_asciicast("spinner-2")
```

To see the full list use the `list_spinners()` function:

```{r list-spinners}
list_spinners()
```

## Showing cli messages in R markdown

Throughout this post I've relied on asciicast to display screencasts of the R console as animated SVG files, rather than what I might normally do and rely on regular R markdown code chunks to do the work. There's a reason for this: the R console is a terminal, and its behaviour doesn't always translate nicely to HTML. Part of the magic of the [rmarkdown](https://rmarkdown.rstudio.com/) package [@rmarkdown] is that *most* of the time it is able to capture terminal output and translate it seamlessly into HTML, and we mere mortal users never notice how clever this is. However, when dealing with cli output, we run into cases where this breaks down and the [law of leaky abstractions](https://www.joelonsoftware.com/2002/11/11/the-law-of-leaky-abstractions/) comes into play: text generated at the R console does not follow the same rules as text inserted into an HTML document, and R Markdown sometimes needs a little help when transforming one to the other.

An important thing to remember about cli is that the text it produces is a *message*,^[Somewhat counterintuitively, although cli emits messages that can be suppressed by `suppressMessages()`, they don't behave *precisely* the same as the messages produced by `message()`. The default handler for base R messages sends the output to the `stderr()` connection and so they are often shown as the dreaded "red text" that users learn to fear. To avoid this, the default behaviour in cli sends messages to the `stdout()` connection, thereby avoiding this issue. However, cli does allow you to control this behaviour: see the `start_app()` and `stop_app()` functions for more information.] so its visibility in R Markdown depends on the chunk option for messages. For this document I set my knitr defaults so that messages and warnings would be shown, but that isn't necessarily the case for every R Markdown template. For this document I included this line in the very first chunk:^[That chunk is not visible in the output because I set `include=FALSE` in the options for that chunk]

```{r knitr-defaults, eval=FALSE}
knitr::opts_chunk$set(message = TRUE, warning = TRUE)
```

This ensures that R markdown will not suppress the cli output, and as a consequence this works:

```{r cli-without-suppression} 
cli_text("I'm still here bitches, and I know everything. -A")
```

However, things are not this simple. Suppose I had used `cli_alert()` in the previous chunk rather than `cli_text()`. Unlike the basic `cli_text()` function, `cli_alert()` uses fancy symbols and colours in the output, and it is at this point that the cracks become exposed. If I had attempted to use `cli_alert()` in the previous chunk, the document would have failed to knit and what I would have gotten is an error message complaining about 

```
PCDATA invalid Char value
``` 

Intuitively you might guess that somewhere in the R Markdown pipeline, an invalid or malformed character has been created.^[As an aside, if you're running a site with an RSS feed it may also write malformed characters into the index.xml file as well as any generated .html file. When I encountered this problem I found that even when I "fixed" my .Rmd file the document wouldn't re-knit, because of the problems with the xml file. Eventually I realised that I could solve the problem by deleting the index.xml file for the RSS feed and *then* knitting again. Sigh] The reason this happens is that the colours and symbols used by cli, and supported in the R console, rely on [ANSI escape codes](https://en.wikipedia.org/wiki/ANSI_escape_code), but those escape codes aren't recognised in HTML and -- apparently -- they can wreak havoc when R markdown writes those characters into the HTML document. ANSI colours in R are usually generated with the help of the [crayon](https://github.com/r-lib/crayon) package [@crayon], and per the [issue #24](https://github.com/r-lib/crayon/issues/24) thread that I encounter on a semi-regular basis, it can be tricky to manage. Solving it requires multiple steps. First, we need to make sure that the R Markdown document turns on crayon support:

```{r crayon-in-rmd}
options(crayon.enabled = TRUE)
```

This isn't the whole solution, however, because all it does is (as I understand it) tell R Markdown to stop ignoring all the ANSI stuff. To get it to handle the ANSI code correctly we need to specify the knit hooks that explicitly tell R Markdown what to do. We can do this with the help of the [fansi](https://github.com/brodieG/fansi) package [@fansi], which contains an obscurely-named function `sgr_to_html()` which translates a subset of the ANSI control sequences to HTML, and ignores all the others. Using this, we can write an `ansi_aware_handler()` function that will take an input string `x` and return HTML output appropriate for the R Markdown context:


```{r ansi-aware-handler}
ansi_aware_handler <- function(x, options) {
  paste0(
    "<pre class=\"r-output\"><code>",
    fansi::sgr_to_html(x = x, warn = FALSE, term.cap = "256"),
    "</code></pre>"
  )
}
```

Next, we tell the [knitr](https://yihui.org/knitr/) package [@knitr] to use this function whenever it needs to handle regular output, and just for good measure we'll do the same for messages, errors, and warnings:

```{r}
knitr::knit_hooks$set(
  output = ansi_aware_handler, 
  message = ansi_aware_handler, 
  warning = ansi_aware_handler,
  error = ansi_aware_handler
)
```

Now that we have done all this obnoxious setup, my R Markdown document renders `cli_alert()` and friends correctly:

```{r cli-alerts}
cli_alert("I'm still here bitches, and I know everything. -A")
```


## Writing tidy messages

Up to this point I've been rather sloppy about how I've constructed my messages. Conceptually, a message should correspond to "one semantically meaningful bundle of information" that might be split over several lines. However, what I've actually been doing with some of the ominous messages is making multiple calls to `cli_text()` that, when pieced together, form a coherent message. To see this, here's a longer message. This is the letter that A sent to Aria's mothe, Ella, informing her of her husband's affair in order to hurt Aria:

```{r tidy-msg-1}
send_cruel_letter_piecewise <- function() {
  cli_text('Your husband, Byron, is involved with another woman')
  cli_text('and when I say involved I mean in a "romantic" way.')
  cli_text('This is not something recent. It started before your')
  cli_text('family went away to Iceland and from the look of')
  cli_text('things, it may be starting up again now that you\'re')
  cli_text('back. I know this is hard to hear, but it is the')
  cli_text('truth. If you don\'t believe this about your husband,')
  cli_text('ask your daughter. She knows all about it.')
  cli_text('Sincerely,')
  cli_text('A')
}

send_cruel_letter_piecewise()
```

This ought to be one message across 10 lines not 10 separate one-line messages, but it's the latter that I've actually implemeneted here. I can fix this by using the `cli()` function, which takes an expression as input and collects all the output into a single message:

```{r tidy-msg-3}
send_cruel_letter_singly <- function() {
  cli({
    cli_text('Your husband, Byron, is involved with another woman')
    cli_text('and when I say involved I mean in a "romantic" way.')
    cli_text('This is not something recent. It started before your')
    cli_text('family went away to Iceland and from the look of')
    cli_text('things, it may be starting up again now that you\'re')
    cli_text('back. I know this is hard to hear, but it is the')
    cli_text('truth. If you don\'t believe this about your husband,')
    cli_text('ask your daughter. She knows all about it.')
    cli_text('Sincerely,')
    cli_text('A')
  })
}

send_cruel_letter_singly()
```

Much nicer. As every would-be tormenter knows, its important to pay attention to the details. 


## Epilogue

```{r epilogue}
cli(
  cli_blockquote(
    quote = "Friends share secrets, that's what keeps us close",
    citation = "Alison"
  )
)
```






